The art of modelling the 3D structure of the genome\\ Insights in {\em P. falciparum} genome organization

### 2017-07-26 doc/manuscript/paper.tex
More on the introduction

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citep{duan:three-dimensional,
burton:species-level, mizuguchi:cohesin-dependent}, bacteria
\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
cell lines \citep{lieberman-aiden:comprehensive, rao:3D}. Within the past few
years, two studies of the three-dimensional structure of {\em P. falciparum}
have been released.  \citet{lemieux:genome-wide} probed several strains of
{\em P. falciparum} to understand the link between 3D structure and the
complex regulation of \textit{var} genes, a family of genes involved the
invasion of red blood cell and also responsible for the parasite's great
capacity to evade our immune system. \citet{ay:three-dimensional} assayed the
3D structure of the genome of the parasite during three key stages of the
erythrocytic cycle. The relatively small size of the genome yet relatively
complicated genome architecture, the complex life cycle and the strong link
between chromatine folding, gene regulation and epigenetic places {\em P.
falciparum} as an interesting case study for developing new methods for better
understanding the properties of genome folding and its link to gene
regulation. An important part of analysing the three-dimensional structure of
the genome of {\em P. falciparum} (as well as many other organisms) relied on
the inferring accurate 3D models of the structure. These models were then used
as a preliminary step to various analysis, such as identifying colocolized
elements, distinguishing between open and closed chromatin and last but not
least, intergrating with other sources of data, such as gene expression or
chromatin modification.

\fixme{Add why we are doing a review}

The study of chromosome organization based on contact count maps broadly
falls into two categories: model-based studies and data-driven studies. The
former methods consider the polymer nature of DNA to leverage the theoretical
and computational work done in statistical physics of polymers to build with
as few assumptions as possible many chromosome conformations. Those chromosome
conformations are then used to compare against experimental data, such as Hi-C
contact count matrices, in order to iteratively improve the models. These
models offer mechanistical insights into the folding of DNA. The latter
approaches use the experimental data to infer 3D models, by typically minizing
a cost function ensuring the models are as consistent as possible with the
data. These data driven models and analysis, with a particular focus on the
one applied to {\em P. falciparum} are the primary focus of this
review.

@@

Though we here review some of the methods used to study and build models, this
is a very incomplete view of a blooming field. \citet{rosa:computational}
provide a more thorough (but again incomplete) overview of computational models
of genome architectures.

>>>>

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citep{duan:three-dimensional,
burton:species-level, mizuguchi:cohesin-dependent}, bacteria
\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
cell lines \citep{lieberman-aiden:comprehensive, rao:3D}. Within the past few
years, two studies of the three-dimensional structure of {\em P. falciparum}
have been released.  \citet{lemieux:genome-wide} probed several strains of
{\em P. falciparum} to understand the link between 3D structure and the
complex regulation of \textit{var} genes, a family of genes involved the
invasion of red blood cell and also responsible for the parasite's great
capacity to evade our immune system. \citet{ay:three-dimensional} assayed the
3D structure of the genome of the parasite during three key stages of the
erythrocytic cycle. The relatively small size of the genome yet relatively
complicated genome architecture, the complex life cycle and the strong link
between chromatin folding, gene regulation and epigenetic places {\em P.
falciparum} as an interesting case study for developing new methods for better
understanding the properties of genome folding and its link to gene
regulation.

An important part of analyzing the three-dimensional structure of the genome
of {\em P. falciparum} (as well as many other organisms) relied on the
inferring accurate 3D models of the structure. These models were then used as
a preliminary step to various analysis, such as identifying colocalized
elements, distinguishing between open and closed chromatin and last but not
least, integrating with other sources of data, such as gene expression or
chromatin modification. In recent years, a profusion of methods were developed
for creating 3D models of the genome, either as standalone methods or with the
specific goal to better understand an organism, a biological process (such as
the inactivation of chromosome X) or the folding of a chromosome. These
methods broadly fall into two categories: ``model-based'' and ``data-driven''.
The former methods consider the polymer nature of DNA to leverage the
theoretical and computational work done in statistical physics of polymers to
build with as few assumptions as possible many chromosome conformations. Those
chromosome conformations are then used to compare against experimental data,
such as Hi-C contact count matrices, in order to iteratively improve the
models. These models offer mechanistical insights into the folding of DNA. The
latter approaches use the experimental data to infer 3D models, by typically
minimizing a cost function ensuring the models are as consistent as possible
with the data.

This paper provides a view on the work accomplished on building and using 3D
models using contact maps for better understanding of genomic and epigenetic
processes, with a particular highlight on the {\em P falciparum}. It dwells on
modeling challenges: why, how and what do 3D models teach us about gene
regulation? The first and second sections of this paper discuss respectively
current data- and model-driven methods to build 3D structures of DNA
using contact maps.  The last section reviews how these models can be used to
discover the key role of genome architecture in biological process, with a
particular focus on discoveries on {\em P. falciparum} genome architecture.

@@
